Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-03 manuscript/manuscript.tex
Begin definitions and non-recombining theory.

@@

\section*{Introduction}
The genetic diversity of a population reflects its demographic and evolutionary history.
Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
The fundamental tools of the trade are toy models, which show how historical forces shape genetic diversity, and which form the basis of parametric inference methods.
However, populations are complicated and, moreover, varied in their complications.
No simple model can capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.

One of the most important and widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002})---tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).

@@

A serious problem for this inference procedure is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For example, \cite{SchriderEtAl2016} recently demonstrated how population size inference can be distorted by selective sweeps.
This effect is present in SFS-based methods as well as in sequentially Markov coalescent methods (e.g., \cite{LiDurbin2011}).
In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce overall genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
Genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of neutral and non-neutral contexts.
Whereas in the Kingman coalescent lineages may only coalesce pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph},
rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.
In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.
It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.

One possible approach to identifying multiple mergers in genomic data to use the SFS as a summary statistic.
(\cite{Koskela2015} developed a full-likelihood method based on importance sampling, but as with other ``exact'' inference methods it does not scale to genomic data.)
\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merge coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.

@@

Here, we propose that summary statistics based on the spectrum of mutation frequencies at pairs of nearby site---the 2-SFS \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
Finally, we demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

\section*{Definitions and single-locus theory}
[Define statistic, show that it works without recombination. (Detects multiple mergers, doesn't respond to population size changes)]

@@

@@

TODO:
- Review and re-write abstract.
-

>>>>

\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\E}[1]{\left< #1 \right>}

@@

\section*{Introduction}
The genetic diversity of a population reflects its demographic and evolutionary history.
Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
The fundamental tools of the trade are toy models, which are used to study how historical forces shape genetic diversity, and which form the basis of parametric inference methods.
However, populations are complicated and, moreover, varied in their complications.
No simple model can capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.

One of the most important and widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
The Kingman coalescent is a stochastic process that generates gene genealogies, trees representing the patterns of shared ancestry of sampled individuals.
Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).

@@

A serious problem for this inference procedure is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For example, \cite{SchriderEtAl2016} recently demonstrated how population-size inference can be distorted by selective sweeps.
This effect is present in SFS-based methods as well as in sequentially Markov coalescent methods (e.g., \cite{LiDurbin2011}).
In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce overall genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
Genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph},
rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.
In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.
It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.

One possible approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
(\cite{Koskela2015} developed a full-likelihood method based on importance sampling, but, as with other ``exact'' inference methods, it does not scale to genomic data.)
\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.

@@

Here, we propose that summary statistics based on the spectrum of mutation frequencies at pairs of nearby sites---the 2-SFS \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
Finally, we demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

\section*{Definitions and non-recombining theory}

\subsection*{The SFS and 2-SFS}
Consider a sample of $n$ chromosomes taking from a population of size $N$ with mutation rate $\mu$ per site per generation.
Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : i \in \{1,\ldots,n-1\} \right\}$, where $\xi_i$ is the fraction of sites at which there is a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}: i \in \{1, \ldots, \floor{n/2} \right\}$. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.

The SFS and folded SFS are single-site statistics.
As such, they can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.
We define the two-site frequency spectrum (2-SFS) at distance $d > 0$,
\begin{align*}
    \left\{ \xi_{ij}(d) : i \in \{1, \ldots, n-1\}, j \in \{i, \ldots, n-1\} \right\},
\end{align*}
as the fraction of pairs of sites separated by $d$ bases with a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011} in a model with selection.
By analogy to the folded SFS, we define the folded 2-SFS, as
\begin{align*}
    \eta_{ij}(d) &= \xi_{ij}(d) \\
                 &+ \delta_{n-2i} \xi_{(n-i),j}(d) \\
                 &+ \delta_{n-2j} \xi_{i,(n-j)}(d) \\
                 &+ \delta_{n-2i} \delta_{n-2j} \xi_{(n-i),(n-j)}(d).
\end{align*}
In this section, we will be considering non-recombining sites and so will suppress the distance $d$.

In the limit of low per-site mutation rate, the expected SFS and 2-SFS are related to moments of the branch length distribution through,
\begin{align}
    \E{\xi_i} &= \mu \E{\tau_i} \\
    \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
\end{align}
where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.

\cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \neq i, (n-i)$.
This result implies a negative association between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \neq i, (n-i)$.
(There are positive associations between mutations at complimentary frequencies induced by trees whose root node partitions the tree into subtrees of size $i$ and $n-i$.)

Later, \cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003} and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \cite{NeherHallatscheck2013}.
Like Fu, Birkner et al. were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}, and thus mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \neq i, n-i$.
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

Our goal is to demonstrate that this distinction in the two-site frequency spectrum between multiple mergers and Kingman coalescent
(1) applies to Kingman coalescents with time-varying coalescent rates,
(2) is robust to recombination between the sites,
(3) is also a feature of forward-time models with selection,
and (4) can form the basis for a model-checking procedure for demographic inference methods.
To this end, we will make use of the fact that the 2-SFS may be seen as a joint probability distribution to define a useful data transformation.

\subsection*{Point-wise mutual information}
\begin{itemize}
    \item the sfs (with zeros) is a probability distribution
    \item the 2-sfs (with zeros) is a joint probability distribution
    \item Mutual information is a functional on joint pmfs that measures the association between two variables $\text{MI}[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)}$
    \item The contributions to MI from each variable are ``pointwise mutual information'' $\text{PMI}(x,y) = \log \frac{p(x,y)}{p(x)p(y)}$.
    \item Thus, we calculate PMI of the expected (folded) 2-SFS as $\text{PMI}(i,j) = \log \left( \frac{\E{\eta_{ij}}}{\E{\eta_i} \E{\eta_j}} \right)$
    \item The advantages of folding are (1) don't need ancestral state and (2) all elements are negative for standard Kingman.
    \item also, note that it doesn't depend on either the mutation rate or the average pairwise coalescent time $T_2$.
\end{itemize}

\subsection*{Population growth vs. the beta coalescent}

\begin{itemize}
    \item Want to calculate PMI for 2-SFS for various models.
    \item implemented Fu and Birkner formulas
    \item \cite{ZivkovicWiehe2008} extended Fu to variable population size.
    \item We implemented their formula for exponential growth and piecewise constant, but in principle could be extended to other models.
    \item Figure 1 shows weighted PMI $\text{wPMI}(i,j) = \E{\eta_{ij}} \text{PMI}(i,j)$.
    \item weighted to highlight regions of the 2-SFS with the most mutations.
    \item Observations:
    \begin{enumerate}
        \item population growth does not qualitatively change the PMI, even though it distorts the SFS a lot.
        \item more multiple mergers leads more positive elements.
        \item Positive associations off-diagonal. Presumably due to uneven branching. Going to ignore because turns out to be sensitive to recombination.
        \item Positive associations between high- and low-frequency mutations. This is the signal we'll focus on.
    \end{enumerate}
    \item Suggests a coarse-graining of the SFS into high- and low-frequency mutations.
    \item Definition of hi-lo corr given cutoff and distance.
\end{itemize}

[Next: Intuition and $n=4$ calculations.]

% \begin{figure}
% \centering
% \includegraphics[width=0.8\textwidth]{figures/trees.png}
% \caption{Possible genealogies with a sample size of $n=4$. Opportunities for singleton/tripleton mutations are in orange. Opportunities for doubletons are in blue.  \label{fig:trees}}
% \end{figure}
%
% We can get an intuitive understanding for this result by considering a sample of size 4. In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
% Furthermore, the total branch length is independent of the topology.
% As a result, there is a trade-off between the branch length leading to singleton/tripleton mutations and the branch length leading to doubletons: loci with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
% Conditional on observing a doubleton at a locus, it is thus more likely that the locus has topology (B) and so the expected number of singletons is lower.
% In terms of the joint site frequency spectrum, we have $\E{\phi_{12}} < \E{\phi_{1}} \E{\phi_{2}}$.
%
% On the other hand, multiple mergers induce correlations between the tree topology and the total branch length.
% For example, topology (C) has less overall opportunity for singletons as well as doubletons than (A) or (B), even though the expected proportion of singletons is higher.
% Thus, observing \emph{any mutation at all} makes topology (C) less likely and the expected number of other mutations at all frequencies more likely.
% If multiple-mergers events are frequent enough, this effect may dominate the tradeoff between (A) and (B) and lead to $\E{\phi_{12}} > \E{\phi_{1}} \E{\phi_{2}}$.
%
% There are two limitations of these existing results.
% First, they only apply to time-homogeneous coalescent processes and thus do not distinguish between multiple-mergers coalescents and Kingman models with time-varying population size.
% Second, they assume a non-recombining locus and may break down for sites separated by a finite genetic distance.

@@

\section*{Methods}
\begin{itemize}
    \item Code to compute moments
    \item Simulation code extending msprime
    \item Forward-time simulation with SLiM
    \item Drosophila
    \begin{itemize}
        \item data pre-processing
        \item fitting demographic model with fastNeutrino
        \item simulating demographic model (could maybe compute exactly)
    \end{itemize}
\end{itemize}

@@
